The brightness and spatial distributions of terrestrial radio sources

### 2013-01-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Slight work on ref II comments
Finalizing referee comments

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{2,3}$,
S.~Zaroubi$^2$,
W.~Brouw$^{2,3}$,
B.~Ciardi$^4$,
I.~T.~Iliev$^5$,
\newauthor
G.~J.~A.~Harker$^6$,
L.~V.~E.~Koopmans$^2$,
G.~Mellema$^7$,
S.~J.~Wijnholds$^3$
\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

\subsection{Spherical case} \label{sec:spherical-geometry}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
\begin{equation}\label{eq:spherical-cap-area}
 A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } a \le R \wedge h \le 2R,
\end{equation}
with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap. Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
\begin{equation} \label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
which equals the formule for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}. Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

\begin{figure}
\begin{center}\hspace{-2mm}\includegraphics[width=8.5cm]{img/plot-rayleigh-and-rfi-combined-trimmed}
\caption{Histograms of simulated samples that all have a contribution of noise and RFI. Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-amplitudes}. Solid lines: the combined distributions, dashed lines: the original distributions before mixing. }
\label{fig:rayleigh-and-rfi-combined}
\end{center}
\end{figure}

@@

We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, also has little effect on the $\log N$ -- $\log S$ histograms. Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, variable $I$ might follow a power-law distribution by itself. It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW. Low-power transmitters, such as remote controls, transmit with an order of 100 mW or even less. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely. Therefore, it is likely that the spatial distribution will dominate the power law in the histogram.

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Name} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

Finally, future RFI excision strategies will further enhance detection accuracy. Currently, RFI excision is applied only on the raw data from single interferometers at high resolution. Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers. We have shown that the current detection algorithms can detect RFI well below the noise. Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable. All RFI that is at that point still below the noise, will act like normal noise and will therefore be harmless.

>>>>

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
A.~G.~de~Bruyn$^{2,3}$,
S.~Zaroubi$^2$%,
%W.~Brouw$^{2,3}$,
%B.~Ciardi$^4$,
%I.~T.~Iliev$^5$,
%\newauthor
%G.~J.~A.~Harker$^6$,
%L.~V.~E.~Koopmans$^2$,
%G.~Mellema$^7$,
%S.~J.~Wijnholds$^3$
\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
%$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
%$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
%$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
%$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
}

@@

\subsection{Spherical case} \label{sec:spherical-geometry}
If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the RFI sources are distributed over the surface of a sphere. If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
\begin{equation}\label{eq:spherical-cap-area}
 A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
\end{equation}
with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap. Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
\begin{equation} \label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
which equals the formule for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}. Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law. Eq.~\eqref{eq:cumulative-spherical} is valid for $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

@@

\begin{figure}
\begin{center}\hspace{-2mm}\includegraphics[width=8.5cm]{img/plot-rayleigh-and-rfi-combined-trimmed}
\caption{Histograms of simulated samples that all have a contribution of noise and RFI. Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-amplitudes}. Solid lines: the combined distributions, dashed lines: the power-law distributions before mixing. }
\label{fig:rayleigh-and-rfi-combined}
\end{center}
\end{figure}

@@

The stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, might also have an effect on the $\log N$ -- $\log S$ histograms. It is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls). Therefore, variable $I$ might follow a power-law distribution by itself. It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well. High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW. Low-power transmitters, such as remote controls, transmit with an order of 100 mW or even less. Therefore, these devices have a spread of around 6 orders of magnitudes in power. As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes. With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely. Therefore, it is likely that the spatial distribution will dominate the power law in the histogram. Otherwise, a turn-over point should be visible in the histogram.

@@

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
 & where SEFD is the System Equivalent Flux Density) & & \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

Finally, future RFI excision strategies will further enhance detection accuracy. Currently, RFI excision is applied only on the raw data from single interferometers at high resolution. Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers. We have shown that the current detection algorithms can detect RFI well below the noise. Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable if it is consistently present. If it is still below the noise and not detectable, but behaves similar to the RFI we are already seeing, it will act like normal noise and will therefore be harmless.
